### 2020-06-18 TCS_version/01-introduction.tex
reviewed directory

Then, we present a methodology for verifying formally
properties of Reaction Systems. The classical notion of bisimulation for 
process algebras allows to consider two processes as equivalent 
where one process can simulate all the actions executed 
by the other one and vice versa. In this paper we define a new
notion of bisimilarity which takes into account the characteristics
of biological systems. We define a new {\em bio-simulation} having in mind
the possibility that two interacting systems may be compared w.r.t.
a subset of the possible biological actions. This is useful for concentrating 
on the sub-model that one may need to consider for a specific study or application,
without getting lost in the complexity of the full biological system, or network.
The notion of bio-simulation relies on a new and simple assertion language,
which allows to focus on some properties of the Reaction Systems to be verified.
In fact, bio-similarity is parametric on a given assertion of interest.
Then we prove that the well-known logical characterisation of bisimilarity in terms of Hennessy-Milner Logic~\cite{HM80} (HML)
can be extended to the case of bio-similarity by tailoring HML formulas to
the same assertion of interest used in bio-similarity.  As HML is a powerful
formalism for specifying properties of labeled transition systems, 
its variant introduced here looks a suitable option to specify formulas 
over complex labels by abstracting from unnecessary details.

@@

A preliminary version of this paper appeared in~\cite{BBF19}.
There are several major differences w.r.t. the paper~\cite{BBF19}, which is here extended as follows:
\begin{itemize}
\item here we present several new examples to illustrate our framework,
and give more detailed explanations about its use;
\item we introduce an assertion language to specify the properties of Reaction Systems;
\item we define the notion of bio-simulation to relate Reaction Systems and compare their behaviours;
\item we show that our assertion language allows to specify properties extending to our
framework the Hennessy-Milner logic;
\item we include here all proofs of main results.
\end{itemize}

>>>>

Then, we present a methodology for verifying formally
properties of Reaction Systems. The classical notion of bisimulation for 
process algebras allows to consider two processes as equivalent 
when one process can simulate all the actions executed 
by the other one and vice versa. In this paper we define a new
notion of bisimilarity which takes into account the characteristics
of biological systems. We define a new {\em bio-simulation} having in mind
the possibility that two interacting systems may be compared w.r.t.
a subset of the possible biological actions. This is useful for concentrating 
on the sub-model that one may need to consider for a specific study or application,
without getting lost in the complexity of the full biological system, or network.
The notion of bio-simulation relies on a new and simple assertion language,
which allows to focus on some properties of the Reaction Systems to be verified.
In fact, bio-similarity is parametric on a given assertion of interest.
Then we prove that the well-known logical characterisation of bisimilarity in terms of Hennessy-Milner Logic~\cite{HM80} (HML)
can be extended to the case of bio-similarity by tailoring HML formulas to
the same assertion of interest used in bio-similarity.  As HML is a powerful
formalism for specifying properties of labeled transition systems, 
its variant introduced here is a suitable option to specify formulas 
over complex labels by abstracting from unnecessary details.

@@

A preliminary version of this paper appeared in~\cite{BBF19}.
There are several major differences w.r.t. the paper~\cite{BBF19}, which is here extended as follows:
\begin{itemize}
\item we present several new examples to illustrate our framework,
and give more detailed explanations about its use;
\item we introduce an assertion language to specify the properties of Reaction Systems;
\item we define the notion of bio-simulation to relate Reaction Systems and compare their behaviours;
\item we show that our assertion language allows to specify properties extending to our
framework the Hennessy-Milner logic;
\item we include here all proofs of main results.
\end{itemize}

### 2020-06-18 TCS_version/02-ReactionSystem.tex
reviewed directory

\begin{definition}[extended interactive process]
Let ${\cal A}=(S,A)$ be a RS, and let $\pi =(\gamma,\delta)$  be an  $n$-step interactive process,
with $ \gamma=\{C_i\}_{i\in[0,n]}$ and  $\delta=\{D_i\}_{i\in[0,n]}$
%, $\delta= D_0, \dots D_n$.
Then, we let $\pi^\infty= (\gamma^\infty,\delta^\infty)$ be the extended interactive process of  $\pi$,
defined as $  \gamma^\infty=\{C_i'\}_{i\in\mathbb{N}}$, $\delta^\infty=\{D_i'\}_{i\in\mathbb{N}}$, where:
$$
C'_j = \left\{
\begin{array}{ll}
C_j & \mbox{if $j\in[0,n]$}\\
\emptyset & \mbox{otherwise ($j>n$)}
\end{array}
\right.
\qquad
D'_j = \left\{
\begin{array}{ll}
D_0 & \mbox{if $j=0$}\\
\mathit{res}_A(D_{j-1}' \cup C_{j-1}') & \mbox{if ($j\geq 1$)}
\end{array}
\right.
$$
% $C_j' = C_j$ for $j\in[0,n]$ and $C_j' =\emptyset$ for $j>n$,   $D_0' = D_0$ and  $D_j' = \mathit{res}_A(D_{j-1}' \cup C_{j-1}')$ for $j\geq 1$.
\end{definition}

>>>>

\begin{definition}[Extended Interactive Process]
Let ${\cal A}=(S,A)$ be a RS, and let $\pi =(\gamma,\delta)$  be an  $n$-step interactive process,
with $ \gamma=\{C_i\}_{i\in[0,n]}$ and  $\delta=\{D_i\}_{i\in[0,n]}$
%, $\delta= D_0, \dots D_n$.
Then, we let $\pi^\infty= (\gamma^\infty,\delta^\infty)$ be the extended interactive process of  $\pi$,
defined as $  \gamma^\infty=\{C_i'\}_{i\in\mathbb{N}}$, $\delta^\infty=\{D_i'\}_{i\in\mathbb{N}}$, where:
$$
C'_j = \left\{
\begin{array}{ll}
C_j & \mbox{if $j\in[0,n]$}\\
\emptyset & \mbox{if $j>n$}
\end{array}
\right.
\qquad
D'_j = \left\{
\begin{array}{ll}
D_0 & \mbox{if $j=0$}\\
\mathit{res}_A(D_{j-1}' \cup C_{j-1}') & \mbox{if ($j\geq 1$)}
\end{array}
\right.
$$
% $C_j' = C_j$ for $j\in[0,n]$ and $C_j' =\emptyset$ for $j>n$,   $D_0' = D_0$ and  $D_j' = \mathit{res}_A(D_{j-1}' \cup C_{j-1}')$ for $j\geq 1$.
\end{definition}
